Hybrid Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-08-13 manuscript/main.tex
score orthogonality proof done

@@

@@

Recalling \eqref{thetaBzetal2i} our goal of showing $

@@

\hat{\boldsymbol{\xi}}_{l_2} = \mathbf{0} \in \mathbb{R}^n.

@@

is equivalent to showing  $	\langle 	\widetilde{\boldsymbol{W}}^{[l_1]}  ,

@@

Using this notation and  \eqref{algorithm_step:residualization_predictor:recall}, we have, with some abuse of notation,

@@

	First note that $\widehat{\boldsymbol{\rho}}_l = \int_{\mathcal{T}} \widetilde{W}^{[l]}[\mathbf{t}] \widehat{\xi}_l[\mathbf{t}] d\mathbf{t}= \widetilde{C}^{*[l]} J^\ast \widehat{\xi}^*_{l}$ for $l=1, \ldots, L$. Then, as above, we first express the $l$-th residualized hybrid predictor in terms of its previous $(l-1)$-th version:
		\widetilde{W}^{[l]}[\mathbf{t}] = \widetilde{W}^{[l]}[\mathbf{t}] - \widehat{\boldsymbol{\rho}}_{l-1} \widehat{\boldsymbol{\delta}}_l[\mathbf{t}]^\top 
	which, after applying the finite-dimensional form (9), implies that
		\setlength{\jot}{10pt}
		\begin{aligned}
			\widetilde{C}^{*[l]}B^*[\mathbf{t}] &=\widetilde{C}^{*[l-1]}B^*[\mathbf{t}] - \widehat{\boldsymbol{\rho}}_{l-1}\widehat{\boldsymbol{\gamma}}^*_{l-1}B^*[\mathbf{t}] \\
			&= \widetilde{C}^{*[l-1]}B^*[\mathbf{t}] - \widehat{\boldsymbol{\rho}}_{l-1} \widehat{\boldsymbol{\rho}}_{l-1}^\top \widetilde{C}^{*[l-1]} J^\ast \left(\widehat{\boldsymbol{\rho}}_{i,l-1}^\top \widehat{\boldsymbol{\rho}}_{l-1} J^\ast \right)^{-1} B^*[\mathbf{t}] \\
			& =\widetilde{C}^{*[l-1]}B^*[\mathbf{t}] -  \frac{\widehat{\boldsymbol{\rho}}_{l-1} \widehat{\boldsymbol{\rho}}_{l-1}^\top \widetilde{C}^{*[l-1]} B^*[\mathbf{t}]}{\widehat{\boldsymbol{\rho}}_{l-1}^\top \widehat{\boldsymbol{\rho}}_{l-1}}  \\
			& =\widetilde{C}^{*[l-1]}B^*[\mathbf{t}] -  \frac{\widetilde{C}^{*[l-1]} J^\ast \widehat{\xi}^{*}_{l-1} \widehat{\boldsymbol{\rho}}_{l-1}^\top \widetilde{C}^{*[l-1]} B^*[\mathbf{t}]}{\widehat{\boldsymbol{\rho}}_{l-1}^\top \widehat{\boldsymbol{\rho}}_{l-1}},
		\end{aligned}
	Since the equation should hold for all argument values $\mathbf{t} \in \mathcal{T}$, we can write:
		\setlength{\jot}{10pt}
		\begin{aligned}
			\widetilde{C}^{*[l]} &= \widetilde{C}^{*[l-1]} -  \frac{\widetilde{C}^{*[l-1]} J^\ast \widehat{\xi}^{*}_{l-1} \widehat{\boldsymbol{\rho}}_{l-1}^\top \widetilde{C}^{*[l-1]}}{\widehat{\boldsymbol{\rho}}_{l-1}^\top \widehat{\boldsymbol{\rho}}_{l-1}} \\
			&= \widetilde{C}^{*[l-1]} \left(I_{MK+p} - \frac{J^\ast \widehat{\xi}^{*}_{l-1} \widehat{\boldsymbol{\rho}}_{l-1}^\top \widetilde{C}^{*[l-1]}}{\widehat{\boldsymbol{\rho}}_{l-1}^\top \widehat{\boldsymbol{\rho}}_{l-1}} \right) \\
			&= \widetilde{C}^{*[j]} \left(I_{MK+p} - \frac{J^\ast \widehat{\xi}^{*}_{j} \widehat{\boldsymbol{\rho}}_{j}^\top \widetilde{C}^{*[j]}}{\widehat{\boldsymbol{\rho}}_{j}^\top \widehat{\boldsymbol{\rho}}_{j}} \right) \widetilde{P},
		\end{aligned}
	where $\widetilde{P}$ is a $(MK+p) \times (MK+p)$ matrix defined as
		\widetilde{P} = \prod \limits_{h=j+1}^{l-1} \left(I_{MK+p} - \frac{J^\ast \hat{\xi}^*_h \widehat{\boldsymbol{\rho}}_{h}^\top \widetilde{C}^{*[h]}}{\widehat{\boldsymbol{\rho}}_{h}^\top \widehat{\boldsymbol{\rho}}_{h}} \right)
	Then, for $j < l$,
		\setlength{\jot}{10pt}
		\begin{aligned}
			\widehat{\boldsymbol{\rho}}_j^\top \widetilde{C}^{*[l]} &=  \widehat{\boldsymbol{\rho}}_j^\top \widetilde{C}^{*[j]} \left(I_{MK+p} - \frac{J^\ast \widehat{\xi}^{*}_{j} \widehat{\boldsymbol{\rho}}_{j}^\top \widetilde{C}^{*[j]}}{\widehat{\boldsymbol{\rho}}_{j}^\top \widehat{\boldsymbol{\rho}}_{j}} \right) \widetilde{P} \\
			&=\left( \widehat{\boldsymbol{\rho}}_j^\top \widetilde{C}^{*[j]} - \frac{ \widehat{\boldsymbol{\rho}}_j^\top \widetilde{C}^{*[j]} J^\ast \widehat{\xi}^{*}_{j} \widehat{\boldsymbol{\rho}}_{j}^\top \widetilde{C}^{*[j]}}{\widehat{\boldsymbol{\rho}}_{j}^\top \widehat{\boldsymbol{\rho}}_{j}} \right) \widetilde{P} \\
			&=\left( \widehat{\boldsymbol{\rho}}_j^\top \widetilde{C}^{*[j]} - \frac{ \widehat{\boldsymbol{\rho}}_j^\top \widehat{\boldsymbol{\rho}}_j  \widehat{\boldsymbol{\rho}}_{j}^\top \widetilde{C}^{*[j]}}{\widehat{\boldsymbol{\rho}}_{j}^\top \widehat{\boldsymbol{\rho}}_{j}} \right) \widetilde{P} \\
			&= \left(  \widehat{\boldsymbol{\rho}}_j^\top \widetilde{C}^{*[j]} -  \widehat{\boldsymbol{\rho}}_j^\top \widetilde{C}^{*[j]} \right) \widetilde{P} \\
			&= \mathbf{0} \in \mathbb{R}^{1 \times (MK+p)},
		\end{aligned}
	Therefore, for $j < l$,
		\widehat{\boldsymbol{\rho}}_j^\top \widehat{\boldsymbol{\rho}}_l =  \widehat{\boldsymbol{\rho}}_j^\top \widetilde{C}^{*[l]} J^\ast \widehat{\boldsymbol{\rho}}_{l} = \mathbf{0} J^\ast \widehat{\boldsymbol{\rho}}_{l}=  0.
	Thus, the proposition is proved.

>>>>

\begin{equation}\label{notation:n_predictors}

@@

\begin{equation}\label{notation:n_inner_products}

@@

Recalling \eqref{thetaBzetal2i},
to achieve our goal of showing $

@@

\hat{\boldsymbol{\xi}}_{l_2} = \mathbf{0} \in \mathbb{R}^n,

@@

it suffices to show
   $	\langle 	\widetilde{\boldsymbol{W}}^{[l_1]}  ,

@@

Using these notations and  \eqref{algorithm_step:residualization_predictor:recall}, we have, with some abuse of notation,
\begin{equation}\label{regression_relation_with_notation_abuse}

@@

			For $l_1 < l_2$,
\widehat{\boldsymbol{\rho}}^{[l_1 ] \top}
\widehat{\boldsymbol{\rho}}^{[l_2]}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\widehat{ \rho }^{[l_1 ]}_i \,
\widehat{ \rho }^{[l_2]}_i
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\widehat{ \rho }^{[l_1 ]}_i \,
\bigl \langle 	\widetilde{W}^{[l_2]}_i  ,
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\big \langle
\widehat{ \rho }^{[l_1 ]}_i \,
 	\widetilde{W}^{[l_2]}_i  ,
\big \rangle_{\mathcal{H}}
Therefore,
to show that $\widehat{\boldsymbol{\rho}}^{[l_1 ] \top}
\widehat{\boldsymbol{\rho}}^{[l_2]} = 0$,
 it suffices to show that
$\sum_{i=1}^n
\widehat{ \rho }^{[l_1 ]}_i \,
\widetilde{W}^{[l_2]}_i  = 0 \in   \mathcal{H}  $,
where this   zero element represents an ordered pair of $K$ zero functions and a zero matrix.
Using Lemma    \ref{proposition:closed_form_orthgonalization}, notations from 
\eqref{notation:n_predictors} and
\eqref{notation:n_inner_products},
and equation \eqref{regression_relation_with_notation_abuse}, we have:
	\widetilde{\boldsymbol{W}}^{[l_1]}  &= 
	\widetilde{\boldsymbol{W}}^{[l_1-1]}
\\&=	\widetilde{\boldsymbol{W}}^{[l_1-1]} - 
	\hat{\xi}^{[l_1-1]}
\widetilde{\boldsymbol{W}}^{[l_1-1]}.
Using this relationship and
with some abuse of notation, we have:
\widehat{ \rho }^{[l_1 ]}_i \,
\widetilde{W}^{[l_2]}_i 
	\widetilde{\boldsymbol{W}}^{[l_1]}
	\hat{\xi}^{[l_1-1]}
	\widetilde{\boldsymbol{W}}^{[l_1-1]}
	\frac{
		\langle 	 	\hat{\boldsymbol{\rho}}^{[l_2] \top}
	\hat{\xi}^{[l_1-1]}
	}{\| 	\hat{\boldsymbol{\rho}}^{[l_1-1]}\|_2^2}
%%%%%%%%%%%%%%%%%%%%%%%%%
	\widetilde{\boldsymbol{W}}^{[l_1-1]}
	\bigr)
\\	&=	h^{[l_1-1]} \bigl( \hat{\boldsymbol{\rho}}^{[l_2] \top}
\widetilde{\boldsymbol{W}}^{[l_1-1]} \bigr)~(\text{say}).
Here, the function $h^{[l_1-1]}: \tilde{ \mathcal{H} } \mapsto \tilde{\mathcal{H}}$ maps the zero element of $\tilde{\mathcal{H}}$ to itself, where the zero element represents an ordered pair of $K$ zero functions and a zero matrix. Using this relationship, we have
	\widetilde{\boldsymbol{W}}^{[l_1]}  
%%%%%%%%%%%%%%%%
	&=	h^{[l_1-1]} \bigl( \hat{\boldsymbol{\rho}}^{[l_2] \top}
	\widetilde{\boldsymbol{W}}^{[l_1-1]} \bigr)
%%%%%%%%%%%%%%%%
h^{[l_1-1]} 
\circ 
h^{[l_1-2]}
\bigl( \hat{\boldsymbol{\rho}}^{[l_2] \top}
\widetilde{\boldsymbol{W}}^{[l_1-2]} \bigr)
%%%%%%%%%%%%%%%%
h^{[l_1-1]} 
\circ 
h^{[l_1-2]}
\circ 
\ldots
\circ 
h^{[l_2]}
\bigl( \hat{\boldsymbol{\rho}}^{[l_2] \top}
\widetilde{\boldsymbol{W}}^{[l_2]} \bigr).
Therefore,
to show that $\widehat{\boldsymbol{\rho}}^{[l_1 ] \top}
\widehat{\boldsymbol{\rho}}^{[l_2]} = 0$,
 it suffices to show $h^{[l_2]}
\bigl( \hat{\boldsymbol{\rho}}^{[l_2] \top}
\widetilde{\boldsymbol{W}}^{[l_2]} \bigr) = 0$:
h^{[l_2]} \bigl( \hat{\boldsymbol{\rho}}^{[l_2] \top}
\widetilde{\boldsymbol{W}}^{[l_2]} \bigr)
	\widetilde{\boldsymbol{W}}^{[l_2]} - 
	\frac{
		\langle 	 	\hat{\boldsymbol{\rho}}^{[l_2] \top}
	}{\| 	\hat{\boldsymbol{\rho}}^{[l_2]}\|_2^2}
	%%%%%%%%%%%%%%%%%%%%%%%%%
	\hat{\boldsymbol{\rho}}^{[l_2] \top
	\widetilde{\boldsymbol{W}}^{[l_2]}
	\bigr)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\widetilde{\boldsymbol{W}}^{[l_2]} - 
\frac{
}{\| 	\hat{\boldsymbol{\rho}}^{[l_2]}\|_2^2}
\hat{\boldsymbol{\rho}}^{[l_2] \top
\widetilde{\boldsymbol{W}}^{[l_2]}
\bigr)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\widetilde{\boldsymbol{W}}^{[l_2]} - 
\frac{
	\hat{\boldsymbol{\rho}}^{[l_2]  } 
}{\| 	\hat{\boldsymbol{\rho}}^{[l_2]}\|_2^2}
\hat{\boldsymbol{\rho}}^{[l_2] \top
\widetilde{\boldsymbol{W}}^{[l_2]}
\bigr)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\widetilde{\boldsymbol{W}}^{[l_2]} 
\hat{\boldsymbol{\rho}}^{[l_2] \top
\widetilde{\boldsymbol{W}}^{[l_2]}
\\&=0.
This completes the proof of Proposition \ref{proposition: orthnormality of PLS scores}.
